### 2018-05-30 submission/manuscript.Rmd
update conclusion and acknowledgements

**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases at 35 cycles (Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Additionally, the median number of OTUs vaired by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2 = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2 = `r min(pull(r2_table, r2))`).

@@

***The bacterial community is similar within polymerase and varies by number of cycles:***
Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 3]. Using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`. For fecal samples, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 at 15 cycles. There was an overall decrease in Bray-Curtis index by cycle comparison group (e.g. the 15 cycle versus 20 cycle group compared to the 30 cycle to 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 3A]. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curits index than the 30 cycle versus 35 cycle comparison group `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05). The mock community has similar trends to the obeservations reported for the fecal bacterial community but none were significant `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B]. Overall, these data suggest that the number of cycles can change the bacterial community independent of the polymerase used to generate the sequences.

***Sequence error varies by polymerase and is consistent across the number of cycles used:***
The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 4 & Table S3]. The majority of the differences across the number of cycles was between Kappa and the other polymerases [Figure 4 and Table S4]. The total sequences with at least one error is also polymerase dependent with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S3 & S4]. These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1]. Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base subsitution.

@@

Standardization of the 16S rRNA gene sequencing workflow may partially solve this problem by introducing a consistent bias to all samples. However, as mentioned in the introduction, many of the choices made to standardize the workflow will result in missing important members of the microbial community. Therefore, the degree of workflow standardization across studies and research groups need to be approached on a study by study basis. All aspects of the 16S rRNA gene sequencing workflow should be customized for the specific scientific questions that wants to be answered. Although a diversity of approaches may make reproducibility and replicability more difficult it will help reduce systematic biases from occuring.

@@

Our observations fill a gap in knowledge on the bias introduced to 16S rRNA gene sequencing results due to differences in polymerases. We found that the number of OTUs and the chimera prevalence is dependent on both polymerase and cycle number chosen. Care should be taken when choosing a polymerase for 16S rRNA gene surveys because their intrinsic differences can change the number of OTUs observed and influence diversity based metrics that do not down-weight rare observations. Knowing the inherent bias associated with different polymerases allows for better interpretation of the relationsip between an individual study to their respective field of research.

@@

The authors would like to thank all the study participants in ERIN whose samples were utilized. We would also like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan. Salary support for Marc A. Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants P30DK034933 and 1R01CA215574.

@@

**Figure 3: Bray-Curtis Community Differences by Five-Cycle Intervals.** A) within person differences based on the next 5-cycle PCR interval in fecal samples. B) Within replicate difference based on the next 5-cycle PCR interval in mock samples. The points represent the median Bray-Curtis index for the samples. The lines represent the range of the minimum and maximum Bray-Curits index value for each PCR 5-cycle increment comparison. The closer a sample is to a Bray-Curtis index of 1.00 the more dissimilar the bacterial community is of the two compared number of cycles.

@@

**Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock Samples.** A) Precentage of chimeric sequences without the removal of chimeras with VSEARCH. C) Percentage of chimeric sequences with the removal of chimeras with VSEARCH. C) The total percent of chimeric sequences removed with VSEARCH by cycle number. The error bars represent the 75% interquartile range of the median.

>>>>

**Results.** When analyzing fecal samples we observed that the range in the number of OTUs detected was not consistent between HiFi DNA polymerases at 35 cycles (Accuprime = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` (min - max) versus Phusion = `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(fecal_numOTU_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Additionally, the median number of OTUs varied by HiFi DNA polymerase used (P-value `r ifelse(max(filter(fs_anova_summary, cycle == "35x")[, "bh"]) < 0.0001, invisible("<"), invisible("check data"))` 0.0001). When analyzing mock community samples the variation in the number of OTUs detected by the polymerases was observable as early as 20 cycles (P-value = `r format(filter(mock_anova_summary, sub_sample_level == 1000, cycle == "20x")[, "bh"], digits = 1, nsmall = 1)`). There also was a large range in the number of OTUs amplified by the polymerases at 35 cycles (Accuprime = `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Accuprime") %>% pull(taq_max)` versus Phusion = `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_min)` - `r filter(mock_num_otu_data, cycles == "35x", taq == "Phusion") %>% pull(taq_max)`). Chimera prevalence in mock communities varied by polymerase with differences being most notable at 35 cycles (Kappa = `r round((filter(error_summary, cycles == "35x", taq == "K") %>% pull(chimera_prevalence_median))*100, digits = 2)`% (median) versus Platinum = `r round((filter(error_summary, cycles == "35x", taq == "PL") %>% pull(chimera_prevalence_median))*100, digits = 2)`%) and this variation persisted after chimera removal using VSEARCH. We also observed positive correlations between chimera prevalence and the number of OTUs with Platinum having the highest (R^2 = `r max(pull(r2_table, r2))`) and Kappa having the worst (R^2 = `r min(pull(r2_table, r2))`).

@@

***The bacterial community is similar within polymerase and varies by number of cycles:***
Within each respective 5-cycle increment comparison there was no difference in Bray-Curtis index between the polymerases (P-value `r ifelse(min(filter(bray_sim_to_prev_cycle_kruskal_results, sub_sample_level == 1000)[, "bh"], na.rm = T) > 0.05, invisible(">"), invisible("check data"))` 0.05) [Figure 3]. Using PERMANOVA to test for community differences based on any of the number of cycles within polymerases, only Phusion had cycle dependent differences (P-value = `r format(filter(bray_permanova_by_taq_results, sub_sample_level == 1000, taq == "PHU")[, "bh"], digits = 1, nsmall = 1)`. For fecal samples, Phusion was one of two polymerases that had enough sequences to be rarefied to 1000 at 15 cycles. There was an overall decrease in Bray-Curtis index by cycle comparison group (e.g. the 15 cycle versus 20 cycle group compared to the 30 cycle to 35 cycle group) (P-value `r ifelse(pull(fecal_kruskal_test_by_cycle_compare_summary, p.value) < 0.01, invisible("<"), invisible(">"))` 0.01) [Figure 3A]. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20 cycle versus 25 cycle comparison groups had a higher Bray-Curtis index than the 30 cycle versus 35 cycle comparison group `r ifelse(min(pull(dunns_test_by_cycle_compare_summary, P.adj)) < 0.05, invisible("<"), invisible(">"))` 0.05). The mock community has similar trends to the observations reported for the fecal bacterial community but none were significant `r ifelse(pull(mock_kruskal_test_by_cycle_compare_summary, p.value) < 0.05, invisible("<"), invisible(">"))` 0.05) [Figure 3B]. Overall, these data suggest that the number of cycles can change the bacterial community independent of the polymerase used to generate the sequences.

***Sequence error varies by polymerase and is consistent across the number of cycles used:***
The median error rate varied by polymerase, with Kappa having the highest error rate of all the polymerases across the number of cycles [Figure 4 & Table S3]. The majority of the differences across the number of cycles was between Kappa and the other polymerases [Figure 4 and Table S4]. The total sequences with at least one error is also polymerase dependent with the majority of differences being between Kappa or Accuprime and the other polymerases [Table S3 & S4]. These differences in error rates were not due to polymerase dependent differences in base substitution rate [Figure S1]. Collectively, the results suggest that sequence error is dependent on polymerase, persists across the number of cycles used, and are not due to any bias towards a specific base substitution.

@@

Standardization of the 16S rRNA gene sequencing workflow may partially solve this problem by introducing a consistent bias to all samples. However, as mentioned in the introduction, many of the choices made to standardize the workflow will result in missing important members of the microbial community. Therefore, the degree of workflow standardization across studies and research groups need to be approached on a study by study basis. All aspects of the 16S rRNA gene sequencing workflow should be customized for the specific scientific questions that wants to be answered. Although a diversity of approaches may make reproducibility and replicability more difficult it will help reduce systematic biases from occurring.

@@

Our observations show that errors dependent on either high fidelity DNA polymerases used or the number of cycles chosen can change diversity metrics used in 16S rRNA gene sequencing results. Care should be taken when choosing a polymerase for 16S rRNA gene surveys because they can influence diversity-based metrics. By knowing the inherent bias associated with different polymerases it allows for more informed choices to be made on how to reduce 16S rRNA gene sequencing-based error.

@@

The authors would like to thank all the study participants in ERIN whose samples were utilized. We would also like to thank Judy Opp and April Cockburn for their effort in sequencing the samples as part of the Microbiome Core Facility at the University of Michigan. Additional thanks to members of the Schloss lab and Dr. Marcy Balunas for reading earlier drafts of the manuscript and providing helpful critiques. Salary support for Marc A. Sze came from the Canadian Institute of Health Research and NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from NIH grants P30DK034933 and 1R01CA215574.

@@

**Figure 3: Bray-Curtis Community Differences by Five-Cycle Intervals.** A) within person differences based on the next 5-cycle PCR interval in fecal samples. B) Within replicate difference based on the next 5-cycle PCR interval in mock samples. The points represent the median Bray-Curtis index for the samples. The lines represent the range of the minimum and maximum Bray-Curtis index value for each PCR 5-cycle increment comparison. The closer a sample is to a Bray-Curtis index of 1.00 the more dissimilar the bacterial community is of the two compared number of cycles.

@@

**Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock Samples.** A) Percentage of chimeric sequences without the removal of chimeras with VSEARCH. C) Percentage of chimeric sequences with the removal of chimeras with VSEARCH. C) The total percent of chimeric sequences removed with VSEARCH by cycle number. The error bars represent the 75% interquartile range of the median.

### 2018-05-30 submission/manuscript.tex
update conclusion and acknowledgements

\textbf{Results.} When analyzing fecal samples we observed that the
range in the number of OTUs detected was not consistent between HiFi DNA
polymerases at 35 cycles (Accuprime = 84 - 106 (min - max) versus
Phusion = 84 - 136). Additionally, the median number of OTUs vaired by
HiFi DNA polymerase used (P-value \textless{} 0.0001). When analyzing
mock community samples the variation in the number of OTUs detected by
the polymerases was observable as early as 20 cycles (P-value = 0.002).
There also was a large range in the number of OTUs amplified by the
polymerases at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
Chimera prevalence in mock communities varied by polymerase with
differences being most notable at 35 cycles (Kappa = 5.71\% (median)
versus Platinum = 26.62\%) and this variation persisted after chimera
removal using VSEARCH. We also observed positive correlations between
chimera prevalence and the number of OTUs with Platinum having the
highest (R\^{}2 = 0.974) and Kappa having the worst (R\^{}2 = 0.478).

@@

\textbf{\emph{The bacterial community is similar within polymerase and
varies by number of cycles:}} Within each respective 5-cycle increment
comparison there was no difference in Bray-Curtis index between the
polymerases (P-value \textgreater{} 0.05) {[}Figure 3{]}. Using
PERMANOVA to test for community differences based on any of the number
of cycles within polymerases, only Phusion had cycle dependent
differences (P-value = 0.03. For fecal samples, Phusion was one of two
polymerases that had enough sequences to be rarefied to 1000 at 15
cycles. There was an overall decrease in Bray-Curtis index by cycle
comparison group (e.g.~the 15 cycle versus 20 cycle group compared to
the 30 cycle to 35 cycle group) (P-value \textless{} 0.01) {[}Figure
3A{]}. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20
cycle versus 25 cycle comparison groups had a higher Bray-Curits index
than the 30 cycle versus 35 cycle comparison group \textless{} 0.05).
The mock community has similar trends to the obeservations reported for
the fecal bacterial community but none were significant \textgreater{}
0.05) {[}Figure 3B{]}. Overall, these data suggest that the number of
cycles can change the bacterial community independent of the polymerase
used to generate the sequences.

\textbf{\emph{Sequence error varies by polymerase and is consistent
across the number of cycles used:}} The median error rate varied by
polymerase, with Kappa having the highest error rate of all the
polymerases across the number of cycles {[}Figure 4 \& Table S3{]}. The
majority of the differences across the number of cycles was between
Kappa and the other polymerases {[}Figure 4 and Table S4{]}. The total
sequences with at least one error is also polymerase dependent with the
majority of differences being between Kappa or Accuprime and the other
polymerases {[}Table S3 \& S4{]}. These differences in error rates were
not due to polymerase dependent differences in base substitution rate
{[}Figure S1{]}. Collectively, the results suggest that sequence error
is dependent on polymerase, persists across the number of cycles used,
and are not due to any bias towards a specific base subsitution.

@@

The authors would like to thank all the study participants in ERIN whose
samples were utilized. We would also like to thank Judy Opp and April
Cockburn for their effort in sequencing the samples as part of the
Microbiome Core Facility at the University of Michigan. Salary support
for Marc A. Sze came from the Canadian Institute of Health Research and
NIH grant UL1TR002240. Salary support for Patrick D. Schloss came from
NIH grants P30DK034933 and 1R01CA215574.

@@

\textbf{Figure 3: Bray-Curtis Community Differences by Five-Cycle
Intervals.} A) within person differences based on the next 5-cycle PCR
interval in fecal samples. B) Within replicate difference based on the
next 5-cycle PCR interval in mock samples. The points represent the
median Bray-Curtis index for the samples. The lines represent the range
of the minimum and maximum Bray-Curits index value for each PCR 5-cycle
increment comparison. The closer a sample is to a Bray-Curtis index of
1.00 the more dissimilar the bacterial community is of the two compared
number of cycles.

@@

\textbf{Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock
Samples.} A) Precentage of chimeric sequences without the removal of
chimeras with VSEARCH. C) Percentage of chimeric sequences with the
removal of chimeras with VSEARCH. C) The total percent of chimeric
sequences removed with VSEARCH by cycle number. The error bars represent
the 75\% interquartile range of the median.

>>>>

\textbf{Results.} When analyzing fecal samples we observed that the
range in the number of OTUs detected was not consistent between HiFi DNA
polymerases at 35 cycles (Accuprime = 84 - 106 (min - max) versus
Phusion = 84 - 136). Additionally, the median number of OTUs varied by
HiFi DNA polymerase used (P-value \textless{} 0.0001). When analyzing
mock community samples the variation in the number of OTUs detected by
the polymerases was observable as early as 20 cycles (P-value = 0.002).
There also was a large range in the number of OTUs amplified by the
polymerases at 35 cycles (Accuprime = 15 - 20 versus Phusion = 14 - 73).
Chimera prevalence in mock communities varied by polymerase with
differences being most notable at 35 cycles (Kappa = 5.71\% (median)
versus Platinum = 26.62\%) and this variation persisted after chimera
removal using VSEARCH. We also observed positive correlations between
chimera prevalence and the number of OTUs with Platinum having the
highest (R\^{}2 = 0.974) and Kappa having the worst (R\^{}2 = 0.478).

@@

\textbf{\emph{The bacterial community is similar within polymerase and
varies by number of cycles:}} Within each respective 5-cycle increment
comparison there was no difference in Bray-Curtis index between the
polymerases (P-value \textgreater{} 0.05) {[}Figure 3{]}. Using
PERMANOVA to test for community differences based on any of the number
of cycles within polymerases, only Phusion had cycle dependent
differences (P-value = 0.03. For fecal samples, Phusion was one of two
polymerases that had enough sequences to be rarefied to 1000 at 15
cycles. There was an overall decrease in Bray-Curtis index by cycle
comparison group (e.g.~the 15 cycle versus 20 cycle group compared to
the 30 cycle to 35 cycle group) (P-value \textless{} 0.01) {[}Figure
3A{]}. Using a Dunn's post-hoc test the 15 cycle versus 20 cycle and 20
cycle versus 25 cycle comparison groups had a higher Bray-Curtis index
than the 30 cycle versus 35 cycle comparison group \textless{} 0.05).
The mock community has similar trends to the observations reported for
the fecal bacterial community but none were significant \textgreater{}
0.05) {[}Figure 3B{]}. Overall, these data suggest that the number of
cycles can change the bacterial community independent of the polymerase
used to generate the sequences.

\textbf{\emph{Sequence error varies by polymerase and is consistent
across the number of cycles used:}} The median error rate varied by
polymerase, with Kappa having the highest error rate of all the
polymerases across the number of cycles {[}Figure 4 \& Table S3{]}. The
majority of the differences across the number of cycles was between
Kappa and the other polymerases {[}Figure 4 and Table S4{]}. The total
sequences with at least one error is also polymerase dependent with the
majority of differences being between Kappa or Accuprime and the other
polymerases {[}Table S3 \& S4{]}. These differences in error rates were
not due to polymerase dependent differences in base substitution rate
{[}Figure S1{]}. Collectively, the results suggest that sequence error
is dependent on polymerase, persists across the number of cycles used,
and are not due to any bias towards a specific base substitution.

@@

The authors would like to thank all the study participants in ERIN whose
samples were utilized. We would also like to thank Judy Opp and April
Cockburn for their effort in sequencing the samples as part of the
Microbiome Core Facility at the University of Michigan. Additional
thanks to members of the Schloss lab and Dr.~Marcy Balunas for reading
earlier drafts of the manuscript and providing helpful critiques. Salary
support for Marc A. Sze came from the Canadian Institute of Health
Research and NIH grant UL1TR002240. Salary support for Patrick D.
Schloss came from NIH grants P30DK034933 and 1R01CA215574.

@@

\textbf{Figure 3: Bray-Curtis Community Differences by Five-Cycle
Intervals.} A) within person differences based on the next 5-cycle PCR
interval in fecal samples. B) Within replicate difference based on the
next 5-cycle PCR interval in mock samples. The points represent the
median Bray-Curtis index for the samples. The lines represent the range
of the minimum and maximum Bray-Curtis index value for each PCR 5-cycle
increment comparison. The closer a sample is to a Bray-Curtis index of
1.00 the more dissimilar the bacterial community is of the two compared
number of cycles.

@@

\textbf{Figure 5: HiFi DNA Polymerase Chimera Prevalence in Mock
Samples.} A) Percentage of chimeric sequences without the removal of
chimeras with VSEARCH. C) Percentage of chimeric sequences with the
removal of chimeras with VSEARCH. C) The total percent of chimeric
sequences removed with VSEARCH by cycle number. The error bars represent
the 75\% interquartile range of the median.
